Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-09-15 manuscript/manuscript.tex
edited section 3

\section{INTRODUCTION} 
\label{sec:intro}

@@

\begin{figure}[tbp]
	\centering
  	\includegraphics[width=0.8\textwidth]{10-year-trend_yearly.pdf}
 	\caption{10-year (2014-2023) annual SR Counts.}
  	\label{fig:10-yr}
\end{figure}

Our investigation focused on a 10-year period (2014--2023).
Figure~\ref{fig:10-yr} shows the annual counts of NYC 311 SR over the
10-year period from 2014 to 2023, revealing a clear upward trend in
public engagement with the system. The most significant increase
occurred in 2020, when the number of requests surged to 3.23 million
due to the COVID-19 pandemic, illustrating the essential role of the
311 system in supporting residents during times of crisis. Overall, SR
counts increased by nearly 49.5\% between 2014 and 2023, reflecting
growing public reliance on the system for addressing non-emergency
issues. The steady rise in SRs can be attributed to the increased
accessibility of the 311 system via online and mobile platforms, as
well as heightened public awareness of the service. While the spike in
2020 was exceptional, the broader trend indicates sustained growth in
service request volume over the decade, highlighting the system's
evolving role in managing both routine city operations and
extraordinary events.

@@

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width=0.8\textwidth]{SRs_by_Agency.pdf}
  	\caption{SR counts by Agency with Cumulative Percentage}
	\label{fig:SRcountbyAgency}
\end{figure}

@@

\begin{figure}[tbp]
 \centering
  \includegraphics[width=0.8\textwidth]{SR_by_Complaint_Type.pdf} 
  \caption{Top 20 complaint type(s) and Cumulative Percentage} 
  \label{fig:SR_complaints}
\end{figure}

@@

Many quality criteria are utilized in order to process high-quality 
data, including:
\begin{comment}
\begin{itemize}
	\item Data validation - this effort can span a number of criteria

	\begin{itemize}
		\item Mandatory fields: Certain data fields cannot be empty.

		\item Data-types: Certain fields must be of the correct type, e.g. numeric, 
		character, date, 5 numeric digits, etc. Typically these are specified 
		in a Data Dictionary.

		\item Domain compliance:  Many data fields must adhere to a specific 
		domain of values, e.g. statuses, state names, zip codes, gender.

	\end{itemize}   
	\item Structural errors to include naming conventions, 
	 fields not in the Data Dictionary, or inconsistent data entry. For example 
	 intermixing blanks, spaces, NA, N/A, and \textless{}NA\textgreater{} 
	 all to indicate the absence of data.

	\item Redundant, unnecessary, or irrelevant fields

	\item Logical inconsistencies such as related fields that violate the 
	nature of that relationship, e.g. a ``due date'' that is before the ``created date''

	\item Accuracy versus precision
\end{itemize}
\end{comment}

This purpose of this analysis effort is to identify the presence of 
errors; not necessarily to correct them. A correction effort should
be undertaken only after an investigation as to the why and how 
such errors came about, and a discussion as to whether or not it 
even is an \textit{error}.

\subsection{Structural Issues}
\label{sec:structural}
Structural issues refers to how data is organized, formatted, 
or presented within a dataset. Structural issues can make 
it difficult to analyze the data effectively. Here are some 
characteristics of the 311 SR data set:

@@

The 311 SR \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}
 identifies 41 data columns (fields) and related information 
 for each column. However, when downloaded the data reveals 
 six additional fields that are not included in the Data Dictionary. 
(Note that these fields do show up on the portal Column Manager widget
 as ``@computed\_region\_xxxx\_xxxx''.) To some extent one can infer 
 what the these @computed\_fields are and how they are determined, but
 not with certainty. Most importantly is the question, can these fields be reliably 
 used for analytical purposes?  These six ``computed'' fields are:

\begin{itemize}
	\item zip\_codes

	\item community\_districts

	\item borough\_boundaries

	\item city\_council\_districts

	\item police\_precincts

	\item police\_precinct 
\end{itemize}

\subsection{Blank \& N/A data}
\label{sec:blanks}
Understanding the absence of data by field is important 
when undertaking analysis. For example, if you wanted to 
determine if SRs were closed before or after their due\_date, 
you would be challenged as 99.6\% of the due\_date field is 
blank. When counting fields for blank or N/A values, they appear 
to divide into three groups: \textbf{Mostly Empty}, \textbf{Partially Empty}, 
and \textbf{Few/None Empty}.

The Mostly Empty category ranges from 93-99.9\% blank 
and includes such fields as taxi\_company due\_date, pickup\_location, and 
landmark. The Partially Empty includes such fields as location\_type, borough, 
and cross\_street. And the Few/None Empty includes created\_date, 
complaint\_type, agency, and status. In some cases it may make 
sense to inquiry as to why some fields are frequently blank. Here is a graphic 
depiction of total empty (blank \& N/As) for each fields illustrating
the grouping into the Most, Partial, and Few categories.

@@

\begin{comment}
 \subsection{Validating Data for Acceptable Values}
 \label{sec:domain}
Any analytic effort needs to be aware of fields which 
contain invalid values. Typically such fields would be
identified and removed from the analysis. Here are the results
of selected field validation.

\begin{itemize}
	\item Latitude and Longitude fields fall within the geographic 
	boundaries of New York City.

	\item The unique\_key field was in-fact unique.

	\item Unfortunately, very few domains of acceptable 
	values are specified in the Data Dictionary. The below 
	fields were tested as complying with their domain as 
	determined by public usage and historical datasets:

	\begin{itemize}
		\item address\_type
		\item status
		\item borough, borough\_boundaries, \& park\_borough 
		\item data\_channel
		\item vehicle\_type
		\item city\_council\_district
	\end{itemize}

\end{itemize}
\end{comment}

\label{sec:zipcodesissues}
 \textbf{Issues with Zip Codes:} Some fields proved to be problematic 
 as to their compliance with a domain of legal values. For example, all zip codes 
 (two fields: zip\_codes and incident\_zip) should be valid as defined 
 by the USPS database containing all 37,946 valid zip codes. However,
 the computed field zip\_codes proved problematic with over 
58\% (3.6 million) of the field entries being invalid. And while the more accurate 
incident\_zip field has only .07\% invalid entries, it still contains
4163 such errors; not insignificant. The breakdown of the invalid 
entries in the zip\_code, sorted by Agency shows that the 
distribution by percentage mirrors the overall breakdown of 
SRs by Agency, potentially indicating a systemic problem.

\label{sec:case-study-zip-codes}
\textbf{Case Study: Noise Complaints by Zip Code.}  Scenario: The NYC 
Office of Nightlife (ONL) wants to know ``What are the top 10 
zip codes for Noise Complaints over the last two years?'' ONL wants
to assess the impact of recent efforts promoting a vibrant nightlife 
in NYC while seeking to ease strained relations between bar 
and club owners. Let's assume that this analysis 
uses the zip\_codes field, one of the computed fields known to 
have validity problems. We then repeat the analysis with the more 
accurate incident\_zip field. Finally, we subject these two analysis effort 
to validation against the USPS zip code database.

@@

As indicated, six out of ten zip\_codes are invalid, which corresponds closely 
with what is observed in the overall dataset (58\% invalid). Whereas 
the incident\_zip field is completely valid, in-line with the 
overall incident\_zip validation percentage of 99\%.

\label{sec:police-precincts}
\textbf{Issues with Police Precincts:} A curious case exists when 
examining two (nearly) identical fields: police\_precincts and 
police\_precinct. Both of those fields are among the computed fields 
previously mentioned. Using 
\href{https://www.nyc.gov/site/nypd/bureaus/patrol/precincts-landing.page}
{NYPD Precinct listings} it is possible to determine the valid 
police precincts. What we find is that both police\_precinct 
and police\_precincts  have about 35\% invalid entries. Unfortunately, 
it's not the same invalid entries. The police\_precincts field, 
has 2,171,864 invalid entries (35\%). Similarly the 
police\_precinct field has 2,171,778 invalid entries 
(also 35\%), but not the same invalid entries.

\label{sec:communityboards}
\textbf{Issues with Community Boards:} Community Boards are the most 
local, grassroots form of City government; a vital connection 
between communities, elected officials, and City agencies. Community 
Boards often service as a measure of the equity of City services 
throughout the five boroughs. In the 2022-2023 dataset there 
are 27,276 invalid community\_board entries representing 0.43\% 
of non-blank data. The distribution of invalid Community Boards 
by Agency is not consistent with the overall SR Agency distribution,
indicating that there are likely issues at key Agencies, in 
this case Taxi \& Limo Commission (TLC), DPR, etc.

\label{sec:communitydistrict}
\textbf{Issues with Community Districts:} Another of the computed fields 
is community\_districts. Community Districts 
are the geographic boundaries for the Community Boards, but differ as they
are used by the Department of City Planning (DCP) for environmental, 
socio-economic, and demographic purposes. It is a geographical 
division rather than a City government division. Due to how the 
community\_district data is formatted, it is not possible to 
establish validity. However, we can determine that the dataset 
contains 72 unique entries, while there are only 59 
valid Community Districts.

\label{sec:negativeduration}
\textbf{Issues with created\_date and closed\_date - Negative Durations:} For 
the created\_date and closed\_date fields, one might 
expect that these two fields are automatically populated by the 
SR application software; associated with perhaps setting an SR status 
to ``new'' or ``closed''. Unfortunately, that does not appear 
to be the case as there are several anomalies with the various date 
fields:

\begin{itemize}
	\item SRs with a closed\_date that occurs before the created\_date.

 	\item created\_date(s) and closed\_date(s) in the far, far distant past.

 	\item created\_date(s) and closed\_date(s) that match to the second.

	\item A large number of SRs closed and/or created exactly at midnight 
	or exactly at noon, to the second.
\end{itemize}

Citizens, NYC Government Officials, and NYC Agencies especially use 
the closed and created dates as a measure of the ``duration'' of the SR 
and hence Agency responsiveness of a service. Duration, while 
not directly present in the dataset, is easily 
computed by (closed\_date - created\_date).

Closed-before-Created: There are 12,251 SRs that are closed before they 
were created, thereby generating a nonsensical ``negative duration''. 
While this is a small percentage overall (0.2\%) these errors can 
have significant impact on response time analysis. Here is a sample:

\begin{table}[tbp]
    \centering
    \caption{Largest Negative Duration errors (days)}
	    \begin{tabular}{l l l r l}
	        \toprule
	        \multicolumn{5}{c}{\textbf{Largest errors (days) excluding 
	        extreme negative values}} \\
	        \midrule
	        \textbf{created\_date} & \textbf{closed\_date} & \textbf{duration} 
	        & \textbf{agency} \\
		        \midrule
		        2023-01-27 14:40:00 & 2022-01-14 14:40:00 & -378.0 & DOT \\
		        2023-01-18 10:06:00 & 2022-01-12 10:06:00 & -371.0 & DOT \\
		        2023-01-27 14:36:00 & 2022-01-22 14:35:00 & -370.0 & DOT \\
		        2023-01-11 11:10:00 & 2022-01-09 11:10:00 & -367.0 & DOT \\
		        2023-12-18 03:13:00 & 2023-01-16 13:10:00 & -335.6 & DOT \\
		        \bottomrule
	    \end{tabular}
    \label{tab:largest-errors}
\end{table}

As noted, the largest errors are shown ``excluding extreme negative values''. We found 
eight SRs with extremely large negative durations (\textless{} -730), all originating
with DHS and all containing an entry of ``1900-01-01'' as the closed\_date. This 
generates extremely large negative durations exceeding -44,601 
days (-122 yrs). Such a large of an anomaly can significantly skew statistical 
results, even though the number of occurrences is small. According, these 
SR rows are removed from the analysis. Excluding these extreme negative
values, an analysis shows the negative-duration issue is a DOT problem, 
where 95\% of these types of errors occur.

This violin chart shows the broad spread of negative-duration SRs, albeit with 
the extremely large negative-duration SR removed. While there are few 
outliers, the magnitude of the negative-durations is troubling and can 
produce bizarre analytical results.

\begin{figure}[tbp]
	 \centering
 	 \includegraphics[width = \textwidth]{negative_duration_SR_violin.pdf}
 \caption{Distribution of Negative Durations}
 \label{fig:negative-duration-violin}
\end{figure}

\label{sec:homlessassistance}
\textbf{Case Study: Homeless Person Assistance.} Scenario: DHS wants to know ``How 
quickly have 311 SRs for ``Homeless Person Assistance'' been resolved 
over the past two years?'' This is a typical request made by 
both the public and City government; analogous to ``How quickly 
is my Agency responding to ``critical'' requests, and does 
that performance vary by Borough, Zip Code, etc. To conduct such an 
analysis, we can use the tools on the Open Data Portal. Here are the steps:

\begin{itemize}
    \item Select created\_date from 2022-2023 and filter by complaint\_type = 
    ``Homeless Person Assistance'' (this yields 55,000 SRs)

    \item Compute the ``duration'' (closed\_date – created\_date)

    \item Take an average of the ``duration'' field  which yields: \textbf{Answer:  -4.8 days}  
\end{itemize}

Clearly that answer is nonsensical. How did such a simple task result 
in an absurd answer? The answer lies in the computation of the ``duration'' 
field. It turns out, there are eight DHS SRs that have a closed\_date 
of ``1900-01-01''. Each of those SRs creates a negative duration of -44,602 
days (-122 years). Those eight SRs are enough to drive the 
average duration of 55,000 Homeless Assistance SRs to a negative 
value; clearly incorrect. In this case, the median is a better measure 
of central tendency. The median is  0.2 days (approx 5 hrs) which will
surely please the DHS.

@@

\label{sec: zeroduration}		
\textbf{Issues with created\_date \& closed\_date - Zero Durations:} A more 
prevalent problem occurs when the closed\_date and created\_date 
are exactly the same -- to the second. Accordingly, this creates 
a \textbf{zero duration'}, again nonsensical. There are 191,141 
such SRs representing 3.1\% of all non-blank data. 99\% of these zero duration 
SRs occur in five Agencies:  (Dept of Mental Health \&Hygiene (DHMH), Dept 
of Transportation (DOT), Dept of Business (DOB), Dept of Sanitation 
(DSNY), and Dept of Environmental Protection (DEP). This Agency distribution
is does not mirror the overall Agency distribution of SRs, likely indicating an 
Agency-specific issue.

@@

\label{sec:midnightandnoon}
\textbf{Issues with created\_date \& closed\_date(s) - Midnight \& Noon:} Another 
problem discovered with the created\_date and 
closed\_date fields - there are an unusually large number of SRs 
created or closed at exactly midnight (00:00:00) and exactly noon (12:00:00), 
to the second. Normal distribution of SR creation and closure 
largely follows the work-day clock with most  SRs created during 
day-light hours, and fewer SRs 	created at night and the early hours 
of morning. However, there are significantly greater numbers 
of SR closed exactly at midnight and noon, as well as a significant 
number of SRs created exactly at noon. Observations include:

\begin{itemize}
	 \item There were 99,779 SRs created exactly at noon (12:00:00)

	 \item There were 235,347 SRs closed exactly at midnight (00:00:00)

	\item There were 105,505 SRs closed exactly at noon. 
\end{itemize}

A unique way to visualize this anomaly is to examine the busiest day 
during this 2-year period (Friday, 2023-09-29). Then aggregate 
the created\_date(s) by minute (with seconds equal to zero), providing
a minute-by-minute look at SR creation on that one day. Note the spike 
at exactly noon (12:00), well beyond the 3\textsigma line.

\begin{figure}[tbp]
	\centering
	\includegraphics[width=\textwidth]
	{2-year-trend_SR_created_by_minute_of_busiest_day.pdf}
	\caption{SRs Created Minute-by-Minute on Busiest Day}
	\label{fig:busiestcreated}
\end{figure}

Repeat the same for the \# of SRs with closed\_date(s) 
on the busiest day (Friday, 2023-09-29). Again, aggregated by 
the minute (with seconds equal to zero). Note the spikes 
at 00:00:00 and 12:00:00 respectively.

\begin{figure}[tbp]
	\centering
	\includegraphics[width=\textwidth]{2-year-trend-SR_closed_by_minute_of_busiest_day.pdf}
	\caption{SRs Closed Minute-by-Minute on Busiest Day}
	\label{fig:busiestclosed}
\end{figure}

>>>>

\section{Introduction} 
\label{sec:intro}

@@

\begin{figure}[tbp]
	\centering
  	\includegraphics[width=0.7\textwidth]{10-year-trend_yearly.pdf}
 	\caption{10-year (2014-2023) annual SR Counts.}
  	\label{fig:10-yr}
\end{figure}

Our investigation focused on a 10-year period (2014--2023), downloaded
on \jy{Give date because the data is updated daily.}
This version of the data is available as a supplement to facilitate
reproducibility of the analysis presented in the sequel.
\jy{How many total? file size? Somewhere we used only two years that
  need to be clarified.}
Figure~\ref{fig:10-yr} shows the annual counts of NYC 311 SR over the
10-year period from 2014 to 2023, revealing a clear upward trend in
public engagement with the system. The most significant increase
occurred in 2020, when the number of requests surged to 3.23 million
due to the COVID-19 pandemic, illustrating the essential role of the
311 system in supporting residents during times of crisis. Overall, SR
counts increased by nearly 49.5\% between 2014 and 2023, reflecting
growing public reliance on the system for addressing non-emergency
issues. The steady rise in SRs can be attributed to the increased
accessibility of the 311 system via online and mobile platforms, as
well as heightened public awareness of the service. While the spike in
2020 was exceptional, the broader trend indicates sustained growth in
service request volume over the decade, highlighting the system's
evolving role in managing both routine city operations and
extraordinary events.

@@

\jy{The plot sizes seem to be fixed in the generation code. That's why
  the fonts are small when scaled. I'd generate them with approxiately
  the expected sizes in the manuscript. I'd also use aspect ratio
  0.618 (golden ratio).
}

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width=0.7\textwidth]{SRs_by_Agency.pdf}
  	\caption{SR counts by Agency with Cumulative Percentage}
	\label{fig:SRcountbyAgency}
\end{figure}

@@

\begin{figure}[tbp]
 \centering
  \includegraphics[width=0.7\textwidth]{SR_by_Complaint_Type.pdf} 
  \caption{Top 20 complaint type(s) and Cumulative Percentage} 
  \label{fig:SR_complaints}
\end{figure}

@@

Many quality criteria are employed to ensure high-quality data 
processing. One of the primary efforts is data validation, which spans 
several critical checks. For instance, mandatory fields must not be 
left empty, ensuring that key information is always captured. 
Additionally, certain fields must conform to specific data types, 
such as numeric, character, or date formats, which are typically 
outlined in a Data Dictionary. Another essential aspect of validation 
is domain compliance, where fields must adhere to a predefined set of 
values, such as statuses, state names, zip codes, or gender. Structural 
errors also play a significant role in data quality, particularly when 
naming conventions or data entries are inconsistent. Common issues 
include fields that do not appear in the Data Dictionary or the 
inconsistent use of blanks, spaces, NA, N/A, or ``<NA>'' to indicate 
missing data. Furthermore, redundant or irrelevant fields can clutter 
datasets, reducing efficiency. Logical inconsistencies are another 
important consideration, such as related fields that violate expected 
relationships, like a ``due date'' that precedes the ``created date.'' 
Lastly, the balance between accuracy and precision is crucial, as both 
must be carefully managed to ensure reliable and meaningful data.

Here we identify the presence of issues in the NYC 311 SR open data;
not necessarily to solve them. A solution effort should
be undertaken only after an investigation as to the why and how 
such issues came about, and a discussion as to whether or not it 
is an \textit{error}.

\subsection{Structural Issues}
\label{sec:structural}

Structural issues refers to how data is organized, formatted, 
or presented within a dataset. Structural issues can make 
it difficult to analyze the data effectively. Here are some 
characteristics of the 311 SR data set:

@@

The 311 SR 
\href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary} 
identifies 41 data columns (fields) and provides related information 
for each. However, upon downloading the data, six additional fields 
appear that are not included in the Data Dictionary. These fields are 
visible in the portal's Column Manager widget, labeled with the prefix 
``@computed\_region.'' The additional fields include \texttt{zip\_codes}, 
\texttt{community\_districts}, \texttt{borough\_boundaries}, 
\texttt{city\_council\_districts}, \texttt{police\_precincts}, and 
\texttt{police\_precinct}. While one can infer the meaning and 
derivation of these computed fields to some extent, uncertainty remains. 
The critical question is whether these fields can be reliably used for 
analytical purposes reliable documentation. In particular, which one
of the two percinct fields should one use? What is the difference
between the computed zip codes and the incident zipcode among the 41
documented variables?

\subsection{Missing Data}
\label{sec:blanks}
Understanding the absence of data by field is important 
when undertaking analysis. For example, if you wanted to 
determine if SRs were closed before or after their
\texttt{due\_date}, you would be challenged as 99.6\% of the
\texttt{due\_date} field is
blank. When counting fields for blank or N/A values, they appear 
to divide into three groups:
\textbf{Mostly Empty}, \textbf{Partially Empty}, 
and \textbf{Few/None Empty}. 
The Mostly Empty category ranges from 93--99.9\% blank 
and includes such fields as
\texttt{taxi\_company due\_date},
\texttt{pickup\_location}, and \texttt{landmark}.
The Partially Empty includes such fields as
\texttt{location\_type}, \texttt{borough}, 
and \texttt{cross\_street}. And the Few/None Empty includes
\texttt{created\_date}, \texttt{complaint\_type},
\texttt{agency}, and \texttt{status}. In some cases it may make
sense to inquiry as to why some fields are frequently blank.
Figure~\ref{fig:blank_fields} presents a graphic 
depiction of total empty (blank \& N/As) for each fields illustrating
the grouping into the Most, Partial, and Few categories.

\jy{Can we lighten the color of the bar to make the pertentages more
  legible?}

@@

\subsection{Validating Data for Acceptable Values}
\label{sec:domain}
Any analytic effort must ensure that fields containing invalid values 
are identified and removed from the analysis. Below are the results of 
selected field validation. The latitude and longitude fields were 
found to fall within the geographic boundaries of New York City, and 
the \texttt{unique\_key} field was indeed unique, as required. 
Unfortunately, the Data Dictionary specifies very few domains of 
acceptable values. However, the following fields were tested and found 
to comply with their expected domains, as determined by public usage 
and historical datasets: \texttt{address\_type}, \texttt{status}, 
\texttt{borough}, \texttt{borough\_boundaries}, \texttt{park\_borough}, 
\texttt{data\_channel}, \texttt{vehicle\_type}, and 
\texttt{city\_council\_district}. Some fields proved to be problematic 
as to their compliance with a domain of legal values.

\paragraph{Zip Codes}
\label{sec:zipcodesissues}
All zip codes (two fields: \texttt{zip\_codes} and \texttt{incident\_zip}) 
should be valid as defined by the USPS database, which contains 
37,946 valid zip codes. However, the computed field \texttt{zip\_codes} 
proved problematic, with over 58\% (3.6 million) of its entries being 
invalid. Although the more accurate \texttt{incident\_zip} field has 
only 0.07\% invalid entries, this still results in 4,163 errors, which 
is not insignificant. The breakdown of invalid entries in the 
\texttt{zip\_codes} field, sorted by Agency, shows that the
\jy{Which agency? Missing a table? We could prepare a supplement where
  we can put additional tables/figures without limit.}
distribution by percentage mirrors the overall breakdown of SRs by 
Agency, potentially indicating a systemic problem.

@@

Consider a case study with the NYC Office of Nightlife (ONL). Suppose 
that ONL wants to know, "What are the top 10 zip codes for Noise 
Complaints over the last two years?" ONL seeks to assess the impact 
of recent efforts promoting a vibrant nightlife in NYC while also 
easing tensions between bar and club owners. Assume this analysis uses 
the \texttt{zip\_codes} field, one of the computed fields known to have 
validity issues. The analysis is then repeated using the more accurate 
\texttt{incident\_zip} field. Finally, both analyses are validated 
against the USPS zip code database. The top 10 zip codes and their 
validation results are summarized in Table~\ref{tab:zipcodes}. Six out 
of the ten \texttt{zip\_codes} entries are invalid, closely matching 
the overall dataset's 58\% invalid rate. In contrast, the 
\texttt{incident\_zip} field is fully valid, consistent with the 
overall \texttt{incident\_zip} validation rate of 99\%.
It is possible that the computed \texttt{zip\_codes} may mean the zip
code of something differnent from the incident. Without proper
documentation, it is confusing to have it in the released data.

\paragraph{Police Precincts}
\label{sec:police-precincts}
A curious case arises when examining two nearly identical fields: 
\texttt{police\_precincts} and \texttt{police\_precinct}. Both fields 
are among the computed fields previously mentioned. Using the 
\href{https://www.nyc.gov/site/nypd/bureaus/patrol/precincts-landing.page}
{NYPD Precinct listings}, it is possible to determine the valid police 
precincts. Upon validation, both \texttt{police\_precincts} and 
\texttt{police\_precinct} contain about 35\% invalid entries. However, 
the invalid entries are not identical. The \texttt{police\_precincts} 
field has 2,171,864 invalid entries (35\%), while the 
\texttt{police\_precinct} field has 2,171,778 invalid entries (also 
35\%), but they do not match.

\paragraph{Community Boards}
\label{sec:communityboards}
Community Boards are the most local, grassroots form of City government, 
serving as a vital connection between communities, elected officials, 
and City agencies. They often act as a measure of the equity of City 
services across the five boroughs. In the 2022-2023 dataset, there are 
27,276 invalid \texttt{community\_board} entries, representing 0.43\% 
of non-blank data. The distribution of these invalid entries by Agency 
is not consistent with the overall SR Agency 
distribution. This suggests that issues may exist within specific key 
Agencies, such as the Taxi \& Limousine Commission (TLC) and the 
Department of Parks and Recreation (DPR).

\paragraph{Community Districts}
\label{sec:communitydistrict}
Another of the computed fields is \texttt{community\_districts}. 
Community Districts define the geographic boundaries for the Community 
Boards but differ in that they are used by the Department of City 
Planning (DCP) for environmental, socio-economic, and demographic 
purposes. They represent a geographical division rather than a City 
government division. Due to the format of the \texttt{community\_district} 
data, it is not possible to directly establish validity. However, the 
dataset contains 72 unique entries, while only 59 valid Community 
Districts exist.

\paragraph{Date and Closed Date}
\label{sec:negativeduration}
For the \texttt{created\_date} and \texttt{closed\_date} fields, one 
might expect that these fields are automatically populated by the SR 
application software when setting an SR status to ``new'' or ``closed.'' 
Unfortunately, this does not seem to be the case, as several anomalies 
exist in the date fields:
\begin{itemize}
    \item SRs with a \texttt{closed\_date} that occurs before the 
    \texttt{created\_date}.
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} in 
    the far distant past.
    \item \texttt{created\_date(s)} and \texttt{closed\_date(s)} that 
    match to the second.
    \item A large number of SRs closed and/or created exactly at midnight 
    or noon, to the second.
\end{itemize}

\subparagraph{Closed before Created: Negative Duration}
Citizens, NYC Government Officials, and Agencies use the created and 
closed dates to measure the ``duration'' of SRs, which reflects Agency 
responsiveness. While duration is not directly present in the dataset, 
it can be easily computed as the difference between
\texttt{(closed\_date} and \texttt{created\_date}.
There are 12,251 SRs where the \texttt{closed\_date} precedes the 
\texttt{created\_date}, generating nonsensical ``negative durations.'' 
Though this represents only 0.2\% of the dataset, these errors can 
significantly impact response time analyses.
Eight SRs with extremely large negative durations ($-$730 days), all 
originating from the Department of Homeless Services (DHS), contain 
an entry of ``1900-01-01'' as the \texttt{closed\_date}. This results 
in negative durations exceeding $-$44,601 days (or 122 years). Such 
anomalies, though rare, can significantly skew statistical results if 
not addressed during data cleansing. As a result, these SR rows are 
removed from the analysis.

\begin{table}[tbp]
  \centering
  \caption{Negative duraton of the largest magnitude (days) excluding
    extreme negative values}
  \begin{tabular}{l l l r l}
    \toprule
    {created\_date} & {closed\_date} & {duration} 
    & \textbf{agency} \\
    \midrule
    2023-01-27 14:40:00 & 2022-01-14 14:40:00 & $-$378.0 & DOT \\
    2023-01-18 10:06:00 & 2022-01-12 10:06:00 & $-$371.0 & DOT \\
    2023-01-27 14:36:00 & 2022-01-22 14:35:00 & $-$370.0 & DOT \\
    2023-01-11 11:10:00 & 2022-01-09 11:10:00 & $-$367.0 & DOT \\
    2023-12-18 03:13:00 & 2023-01-16 13:10:00 & $-$335.6 & DOT \\
    \bottomrule
  \end{tabular}
  \label{tab:largest-errors}
\end{table}

\begin{figure}[tbp]
	 \centering
 	 \includegraphics[width = 0.7\textwidth]{negative_duration_SR_violin.pdf}
 \caption{Distribution of Negative Durations}
 \label{fig:negative-duration-violin}
\end{figure}

Excluding the extreme negative values, 
Figure~\ref{fig:negative-duration-violin} shows the broad spread of 
negative-duration SRs. While there are few outliers, the magnitude 
of the negative durations is troubling and can produce bizarre 
analytical results. Table~\ref{tab:largest-errors} summarizes the 
negative durations of the largest magnitude (in days). The data 
suggests that the negative duration issue is predominantly a problem 
within the Department of Transportation (DOT), where 95\% of these 
errors occur.

% For illustration, consider a case study of homeless person
% assistance. Suppose that DHS wants to know ``How 
% quickly have 311 SRs for ``Homeless Person Assistance'' been resolved 
% over the past two years?'' This is a typical request made by 
% both the public and City government; analogous to ``How quickly 
% is my Agency responding to ``critical'' requests, and does 
% that performance vary by Borough, Zip Code, etc. To conduct such an 
% analysis, we can use the tools on the Open Data Portal. Here are the steps:

% \begin{itemize}
%     \item Select created\_date from 2022-2023 and filter by complaint\_type = 
%     ``Homeless Person Assistance'' (this yields 55,000 SRs)

%     \item Compute the ``duration'' (closed\_date – created\_date)

%     \item Take an average of the ``duration'' field  which yields: \textbf{Answer:  -4.8 days}  
% \end{itemize}

% Clearly that answer is nonsensical. How did such a simple task result 
% in an absurd answer? The answer lies in the computation of the ``duration'' 
% field. It turns out, there are eight DHS SRs that have a closed\_date 
% of ``1900-01-01''. Each of those SRs creates a negative duration of -44,602 
% days (-122 years). Those eight SRs are enough to drive the 
% average duration of 55,000 Homeless Assistance SRs to a negative 
% value; clearly incorrect. In this case, the median is a better measure 
% of central tendency. The median is  0.2 days (approx 5 hrs) which will
% surely please the DHS.

@@

\subparagraph{Resolution Action Update Date}

\begin{figure}[tbp]
  \centering
  \includegraphics[width = 0.7\textwidth]{post_closed_violin_chart.pdf}
  \caption{Post-closed resolution action update date(s) over 30 days.}
  \label{fig:resolution-violin}
\end{figure}

\jy{Are these from 2-year or 10-year period? Need to clarify in
  Section 2.}

A similar issue occurs when resolution action update date is after the
closing date. \jy{It might be possible, I think, if they allow updates
  after a SR is closed.}
When an SR is updated, the 311 software automatically populates the 
\texttt{resolution\_action\_update\_date}. Some of these updates 
occur long after the SR is closed. As shown in Figure~\ref{fig:resolution-violin}, 
there are 7,460 SRs that were updated more than 30 days after the 
\texttt{closed\_date}, but less than 730 days, to exclude infeasible 
dates like 1900-01-01. The chart highlights the distribution of these 
post-closure updates, with a notable concentration of updates within 
the 30-to-90-day range. This pattern raises questions about whether 
such delayed updates are standard operational practice or indicative 
of a potential issue requiring further investigation. While the reason 
for these delayed updates is unclear, they may impact the accuracy of 
response time analyses if not properly accounted for.

\subparagraph{Same Created Date \& Closed Date: Zero Durations}
A more prevalent issue occurs when the \texttt{closed\_date} and 
\texttt{created\_date} are exactly the same, down to the second. This 
creates a \textbf{zero duration}, which is also nonsensical. There 
are 191,141 such SRs, representing 3.1\% of all non-blank data. 
Ninety-nine percent of these zero-duration SRs occur in five agencies:
DHMH, DOT, DOB, DSNY, and DEP.. 
This distribution does not mirror the overall SR distribution by 
agency, suggesting an agency-specific issue.

@@

\subparagraph{Created Date or Closed Date(s) at Midnight or Noon}
Another issue found with the \texttt{created\_date} and 
\texttt{closed\_date} fields is the unusually large number of SRs 
created or closed at exactly midnight (00:00:00) or noon (12:00:00), 
down to the second. Normally, SR creation and closure follow the 
work-day schedule, with most SRs created during daylight hours and 
fewer at night or in the early morning. However, there is a 
significantly higher number of SRs closed exactly at midnight and noon, 
as well as a large number created exactly at noon. Specific observations 
include:
\begin{itemize}
    \item 99,779 SRs were created exactly at noon (12:00:00).
    \item 235,347 SRs were closed exactly at midnight (00:00:00).
    \item 105,505 SRs were closed exactly at noon.
\end{itemize}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=0.7\textwidth]{2-year-trend_SR_created_by_minute_of_busiest_day.pdf}
  \includegraphics[width=0.7\textwidth]{2-year-trend-SR_closed_by_minute_of_busiest_day.pdf}
  \caption{SRs Created and closed minute-by-minute on busiest day
    during two year period \jy{give years}.}
  \label{fig:busiest}
\end{figure}

Figure~\ref{fig:busiest} visualizes this anomaly by examining the 
busiest day during this 2-year period (Friday, 2023-09-29). The 
\texttt{created\_date} values are aggregated by minute (with seconds 
equal to zero), providing a minute-by-minute look at SR creation on 
that day. Notably, there is a spike exactly at noon (12:00), well 
beyond the 3\textsigma{} line. \jy{May not want to mention 3 sigma as
  here we have highly skewed count data.}
The same analysis was repeated for 
\texttt{closed\_date} values on the busiest day, aggregated by minute 
(with seconds equal to zero). Spikes were observed at both 00:00:00 
and 12:00:00.

These unusual patterns of SR creation and closure exactly at midnight 
and noon suggest the presence of a bulk create/close software process 
that automatically assigns time stamps of midnight (00:00:00) or noon 
(12:00:00) to large batches of SRs. This behavior distorts the 
calculated duration of these SRs by providing inaccurate dates. The 
distribution by agency shows that over 90\% of the ``created-at-noon'' 
SRs come from just two agencies: DOB and DSNY. Further investigation 
reveals that a single agency, DSNY, is responsible for over 99\% of 
the ``closed-at-noon'' SRs.

\subsection{Accuracy and precision}
\label{sec:precision}

\jy{This is caused by reading the number in as double
  precision. Single precision would be good.}

An question of precision vs. accuracy arises with the Latitude 
and Longitude fields. Both Latitude are expressed as 
a 14-decimal number, e.g. 40.86769186022511. Given 
that 1 degree of latitude at the equator is equal to 111.044736 
kilometers, the ``1'' at the end of that number represents 
approximately 1.1 nanometers (1/1,000,000,000 of a meter). For 
reference a DNA molecule is approximately 2nm in width. Clearly 
the representation of the Latitude and Longitude fields is a 
classic case of 14-digit precision, but limited accuracy.

\subsection{Redundant \& Duplicate fields}
\label{sec:duplicates}

During this analysis, several redundant fields were observed and should 
be examined for possible consolidation.
